---
title: "Euler plot"
author: "Mia"
date: "2026-01-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#Libraries

```{r}

library(eulerr)
library(tidyverse)
library(dplyr)
library(readxl)

```


#Data import

Pulling data from the gene list page that also has all phenotypes listed (*HPGD Non-analgesic genes_2025-08-26.xlsx, sheet Real Gene List w Phenotypes*).

The below code can probably be cleaned up to be more generizable/less location dependent, but that is a project for another day.

```{r}

data <- read_excel("Z:/Morphine studies/2024 CCS Study/Mia/Gene list/HPGD Non-analgesic genes_2025-08-26.xlsx", sheet = "Real Gene List w Phenotypes")

```


First time using the *eulerr* package. Attempting to do any manipulations necessary to make it work.

Subset/manipulate data to only have relevant variables

```{r}

euler_data <- data %>%
  rename(Genes = "Gene name") %>%
  subset(select = c(Genes, Phenotypes)) %>%
  mutate(Phenotypes = str_replace_all(Phenotypes, ", ", "&"), 
         Phenotypes = str_replace_all(Phenotypes, " ", "_"), 
         Phenotypes = str_replace_all(Phenotypes, "pain", "Pain"), 
         Phenotypes = str_replace_all(Phenotypes, "Pain,", "Pain"))

```

Pivoting out *euler_data* so each phenotype has it's own column with a value of either TRUE or FALSE for each gene.

```{r}

pivot_data <- euler_data %>%
  separate_longer_delim(Phenotypes, "&")

```

Manipulating data into a form that can be parsed by *eulerr*
- Phenotypes condensed into 4 categories
  - Analgesia
  - Clinical Pain Conditions
  - Nociception
  - Post-operative Pain
- Rows collapsed so that each gene has one row, with all associated phenotypes in that row

A new dataset is created with the counts of each phenotype and phenotype combination listed. This is then pivoted so each phenotype/phenotype combo is a column, with one row each, containing the number of times it was seen in the original dataset.

```{r}

condensed_data <- pivot_data %>%
  mutate(Phenotypes = str_replace_all(Phenotypes, "Migraine", "Clinical_Pain_Conditions"),
         Phenotypes = str_replace_all(Phenotypes, "Arthritis", "Clinical_Pain_Conditions"),
         Phenotypes = str_replace_all(Phenotypes, "Cancer_Pain", "Clinical_Pain_Conditions"),
         Phenotypes = str_replace_all(Phenotypes, "Fibromyalgia", "Clinical_Pain_Conditions"),
         Phenotypes = str_replace_all(Phenotypes, "Musculoskeletal_Pain", "Clinical_Pain_Conditions"),
         Phenotypes = str_replace_all(Phenotypes, "Neuraxial_Pain", "Clinical_Pain_Conditions"),
         Phenotypes = str_replace_all(Phenotypes, "Neuropathic_Pain", "Clinical_Pain_Conditions"),
         Phenotypes = str_replace_all(Phenotypes, "Other_Clinical_Pain", "Clinical_Pain_Conditions"),
         Phenotypes = str_replace_all(Phenotypes, "Temporomandibular_Disorder", "Clinical_Pain_Conditions"),) %>%
  group_by(Genes) %>%
  distinct(Phenotypes, .keep_all = TRUE) %>%
  summarize(Phenotypes = paste(Phenotypes, collapse = "&"))

condensed_count <- condensed_data %>%
  count(Phenotypes) %>%
  pivot_wider(names_from = Phenotypes, values_from = n)

```

Coercing data into euler format

```{r}

panel_a <- euler(setNames(as.numeric(condensed_count[1, ]), names(condensed_count)), shape = "ellipse")

```

Making the plot

```{r}

blobs <- plot(panel_a, quantities = list(fontsize = 9),
     fills = list(fill = c("#FBB4AE", "#B3CDE3", "#CCEBC5", "#DECBE4"), alpha = 0.7),
     edges = "black",
     main = list(label = "Phenotype Categories of Pain-Related Genes", cex = 0.9),
     # quantities = TRUE,
     legend = list(labels = c("Analgesia", "Clinical Pain Conditions*", "Nociception", "Post-operative Pain"), side = "right"))

```
