---
title: "Forest Plots"
author: "Mia Simmons"
date: "2025-05-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This file will include all the code for creating the forest plot. It will turn the (very large) file containing the results of the logistic regression genetic analyses into said plots.

------------------------------------------------------------------------

# Setup

## Libraries

These are the libraries that must be loaded to run this code. Edit as needed if a library must be installed.

```{r}
library(tidyverse)
library(dplyr)
library(here)
library(gt)
library(patchwork)
library(readxl)
```

## Data

The data was loaded in from an Excel file using the *readxl* and *here* packages. The table is saved in R as *VEPORtable_forest*, same as the table the data is pulled from.

```{r}

# import data

VEPORtable_forest <- read_excel(here("filname.xlsx"))

```

------------------------------------------------------------------------

## Data Manipulation

This section will include all coding for the manipulation of *VEPORtable_excel*. The goal of this section is to remove unneccessary columns and rows. To be removed:

-   Columns that were from the unadjusted linear regression (Columns E-H)

-   Rows that do not contain:

    -   p-values \< 0.05

    -   CADD scores \> 10
    
    -   And are not missense variants

The final table will be named *filtered_variants*

```{r}

# drops the columns from the unadjusted linear regression

adjusted_reg <- subset(VEPORtable_forest, select = -c(OR, L95...6, U95...7, P...8))

glimpse(adjusted_reg)

```

```{r}

# renames the columns to be more functional

adjusted_reg <- adjusted_reg %>%
  rename(OR = Or, L95 = L95...10, U95 = U95...11, P = P...12, CADD_score = CADD_PHRED)

```

```{r}

# combines the Gene column with the onlyrsID column

adjusted_reg <- adjusted_reg %>%
  unite(Gene_variant, c(Gene, onlyrsID), sep = " ", remove = FALSE)

```

```{r}

# removes rows with P-values > 0.05, CADD scores < 10

filtered_variants <- adjusted_reg %>%
  filter(P <= 0.05 & CADD_score >= 10 & Consequence == "missense_variant")

```

This is all the manipluation required for this dataset

---

## Making the Forest Plot

This forest plot will be made based of the tutorial on https://www.khstats.com/blog/forest-plots/

### Forest Plot

The main part of the figure, containing the most vital information.

```{r}

# creates the forest plot, the main part of the figure

forest <- filtered_variants %>%
  ggplot(aes(y = fct_rev(Gene_variant))) +
  geom_point(aes(x = OR), shape = 15, size = 1.5) +
  geom_linerange(aes(xmin = L95, xmax = U95)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  labs(x = "Odds Ratio", y = "") +
  coord_cartesian(ylim = c(1,4)) +
  scale_x_continuous(breaks = seq(0,12, by = 1)) +
  theme_classic() +
  theme(axis.line.y = element_blank(),
        axis.ticks.y= element_blank(),
        axis.text.y= element_blank(),
        axis.title.y= element_blank())

forest

```

### List of OR and Confidence intervals

```{r}

# Reformat data for left and right sides of plot

rounded_data <- filtered_variants %>%
  # round OR, 95% CIs, and CADD score to 2 decimal places for journal specifications
  mutate(
    across(
      c(OR, L95, U95, CADD_score),
      ~str_pad(
        round(
          .x, 2), width = 4, pad = "0", side = "right"
        )
      ),
    # add an "-" between CIs
    confi_inter = paste0(
      OR, " (", L95, "-", U95, ")"
      )
    ) %>%
  # round p-values to two decimal places, except in cases where p < .001
  mutate(
    P = case_when(
      P < 0.001 ~ "<0.001",
      round(
        P, 2) == 0.05 ~ as.character(
          round(
            P, 3)
          ),
    P < .01 ~ str_pad( # if less than .01, go one more decimal place
      as.character(
        round(P, 3)),
      width = 4, pad = "0", side = "right"
        ),
    TRUE ~ str_pad( # otherwise just round to 2 decimal places and pad string so that .2 reads as 0.20
      as.character(round(P, 2)),
      width = 4, pad = "0", side = "right"
        )
      )
    ) %>%
  # add a row of data that are actually column names which will be shown on the plot in the next step
  bind_rows(
    data.frame(
      Gene_variant = "Gene Variant",
      confi_inter = "Odds Ratio (95% CI)",
      L95 = "",
      U95 = "",
      P = "p-value",
      CADD_score = "CADD score"
    )
  ) %>%
  # trying to get the order of things to match up
  mutate(Gene_variant = fct_rev(fct_relevel(Gene_variant, "Gene Variant")))

```

```{r}

# Show ORs and CIs for each of the 3 variants

p_left <- rounded_data %>%
  ggplot(aes(y = Gene_variant)) +
  geom_text(aes(x = 0, label = Gene_variant), hjust = 0, fontface = ifelse(rounded_data$Gene_variant == "Gene Variant", "bold", "bold.italic"), size = ifelse(rounded_data$Gene_variant == "Gene Variant", 4, 3.5)) +
  geom_text(aes(x = 1, label = confi_inter), hjust = 0, fontface = ifelse(rounded_data$confi_inter == "Odds Ratio (95% CI)", "bold", "plain"), size = ifelse(rounded_data$confi_inter == "Odds Ratio (95% CI)", 4, 3.5)) +
  geom_text(aes(x = 2, label = CADD_score), hjust = 0, fontface =
ifelse(rounded_data$CADD_score == "CADD score", "bold", "plain"), size = ifelse(rounded_data$CADD_score == "CADD score", 4, 3.5)) +
  coord_cartesian(xlim = c(0, 4)) +
  theme_void()

p_left

```

### List of p-values

```{r}

# Show p-values for each of the 3 variants

p_right <- rounded_data %>%
  ggplot() +
  geom_text(
    aes(x = 0, y = Gene_variant, label = P),
    hjust = 0,
    fontface = ifelse(rounded_data$P == "p-value", "bold", "plain"), 
    size = ifelse(rounded_data$P== "p-value", 4, 3.5)
  ) +
  theme_void() 

p_right

```

### Whole plot

```{r}
# Puts the left, middle, and right parts of the plot together

layout <- c(
  area(t = 0, l = 0, b = 30, r = 25), # left plot, starts at the top of the page (0) and goes 30 units down and 20 units to the right
  area(t = 1, l = 17, b = 30, r = 35), # middle plot starts a little lower (t=1) because there's no title. starts at l=11, goes to the bottom of the page (30 units), and 7 units further over from the left plot (r=18)
  area(t = 0, l = 35, b = 30, r = 38) # right most plot starts at top of page, begins where middle plot ends (l=18, and middle plot is r=18), goes to bottom of page (b=30), and extends two units wide (r=20)
)

# final plot arrangement
p_left + forest + p_right + plot_layout(design = layout)

```

##Exporting plot

Now that the graph has been made, it needs to be exported.

```{r}
# Save figure as a png

#ggsave("APT Forest Plot.png", width = 9.5, height = 5)

#ggsave("CSPT Forest Plot.png", width = 11, height = 5)

```


