---
title: "Gene-snp-assoc"
author: "Mia Simmons"
output: html_document
date: "2025-09-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Setup

Packages needed

```{r}
library(tidyverse)
library(dplyr)
library(here)
library(readxl)
```

#Importing required datasets

The data was loaded in from an Excel files using the *readxl* and *here* packages. The tables are saved in R as *Adjusted_plink* and *Unadjusted Plink*, from the Adjusted and and Unadjusted logistic regression analyses for Objective 1, respectively.

```{r}

Adjusted_plink <- read_excel(here("Objective 1 - Adjusted Logistic Regression Plink_2025-08-28.xlsx"))

Unadjusted_plink <- read_excel(here("Objective 1 - Unadjusted Logistic Regression Plink_2025-08-28.xlsx"))

Gene_list <- read_excel(here("Slim Pain Gene List.xlsx"))

```


#Reformatting datasets

```{r}

Format_adj_plink <- subset(Adjusted_plink, select = -c(MARKER, TEST, NMISS, SE, STAT, TRAIT, JOBID)) %>%
  #rename(OR = ORX) %>%
  rename(Chromosome = CHR)

Format_unadj_plink <- subset(Unadjusted_plink, select = -c(MARKER, TEST, NMISS, SE, STAT, TRAIT, JOBID)) %>%
  #rename(OR = ORX) %>%
  rename(Chromosome = CHR)

Gene_list <- Gene_list %>%
  #rename(Chromosome = Chromsome)  %>%
  rename(Start = "Start -500") %>%
  rename(End = "End +500")

```


#Combining datasets

Where DF1 is your SNP list (containing an unclassified gene, snp location, and chromosome) and DF2 is your genes list (containing a classified gene, chromosome, start and end locations).

combined <- DF2[DF1, 
              on = .(Chromosome, Start <= SNP, End >= SNP), 
              nomatch = 0L,
              .(Name, Chromosome, SNP, Gene)]

combined

```{r}

# combined <- Gene_list[Format_adj_plink,
#                       on = .(CHR == Chromosome), 
#                       .(SNP, CHR, BP, Gene = i.Gene),
#                       nomatch = 0L]
#                       [BP >= Start & BP <= End]

# merged <- merge(Gene_list, Format_adj_plink, by.x = "CHR", by.y = "Chromosome", all.x = TRUE)

# require(data.table)
# 
# genes_start <- as.data.table(genes)
# ## create the start bound as a separate column to join to
# genes_start[,`:=`(start_bound = start - 10)]
# setkey(genes_start, chromosome, start_bound)
# 
# markers <- as.data.table(markers)
# setkey(markers, chromosome, position)
# 
# new <- genes_start[
#     ##join genes to markers
#     markers, 
#     ##rolling the last key column of genes_start (start_bound) forward
#     ##to match the last key column of markers (position)
#     roll = Inf, 
#     ##inner join
#     nomatch = 0
# ##rolling join leaves positions column from markers
# ##with the column name from genes_start (start_bound)
# ##now vector scan to fulfill the other criterion
# ][start_bound <= end + 10]
# ##change names and column order to match desired result in question
# setnames(new,"start_bound","position")
# setcolorder(new,c("chromosome","gene","start","end","marker","position"))
#    # chromosome gene start end marker position

require(data.table)

genes_start <- as.data.table(Gene_list)
## create the start bound as a separate column to join to
genes_start[,`:=`(start_bound = Start - 10)]
setkey(genes_start, Chromosome, start_bound)

as.character(as.numeric(Format_adj_plink$Chromosome))

markers <- as.data.table(Format_adj_plink)
setkey(Chromosome, SNP, BP)

new <- genes_start[
    ##join genes to markers
    markers, 
    ##rolling the last key column of genes_start (start_bound) forward
    ##to match the last key column of markers (position)
    roll = Inf, 
    ##inner join
    nomatch = 0
##rolling join leaves positions column from markers
##with the column name from genes_start (start_bound)
##now vector scan to fulfill the other criterion
][start_bound <= End + 10]
##change names and column order to match desired result in question
setnames(new,"start_bound","BP")
setcolorder(new,c("Chromosome","Gene","Start","End","SNP","BP"))

combined

```

#Print combined table to Excel format
