---
title: "Gene-snp-assoc"
author: "Mia Simmons"
output: html_document
date: "2025-09-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Setup

Packages needed

```{r}
library(tidyverse)
library(dplyr)
library(readxl)
library(writexl)
library(here)
library(data.table)
library(R.utils)
```

#Importing and reformatting required datasets

Each chunk loads in a file.

##Genes

```{r}

genes <- read.csv("location/filename.csv")

```

##PLINK Data

```{r}

# plink <- fread("Flocation/filename plink.assoc.logistic", sep = "auto")



```


#Reformatting datasets

Removes and renames columns as necessary. Modify to fit dataset.

##Genes

```{r}

slim_genes <- genes %>%
  rename(Start = Start..500, End = End..500, Gene = Gene.name) %>%
  filter(Chromosome != "X") %>%
  distinct(Gene, .keep_all = TRUE) %>%
  mutate(Chromosome = as.numeric(Chromosome))
  #group_by(Chromosome) #didn't end up using this

```


##PLINK

```{r}

slim_plink <- plink %>%
  subset(select = -c(MARKER, TEST, NMISS, SE, STAT, TRAIT, JOBID)) %>%
  rename(Chromosome = CHR, OR = ORX) %>%
  group_by(Chromosome)

```


#Combining datasets

Much simpler than I assumed, unless I'm missing something. Defines first where the join should occur (join at chromosomes where the BP of the first dataset is between the start and end columns of the second dataset), then full joins the first dataset (plink data) with the second dataset (gene list), then views the resulting table

##Adjusted

```{r}

by <- join_by(Chromosome, between (BP, Start, End))

plink_summ <- full_join(slim_plink, slim_genes, by)

view(plink_summ)

```


#Print combined table to excel format

Saves the dataset to working directory

```{r}

# write_xlsx(plink_summ,"DESIRED FILE NAME.xlsx")


```

