---
title: "Gene blobs"
author: "Mia"
date: "2025-10-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Introduction

The purpose of this code is to create a graph of the genes in my gene list from HPGDB, with overlapping groups showing the phenotypes. An attempt will be made to show all gene names; if unsuccessful, the number of genes in each category will be shown instead. It would be nice to represent the chromosomes each gene is on in some way, but this may not be possible.

#Libraries

```{r}

library(tidyverse)
library(dplyr)
library(readxl)

```


#Data import

Pulling data from the gene list page that also has all phenotypes listed (*HPGD Non-analgesic genes_2025-08-26.xlsx, sheet Real Gene List w Phenotypes*).

The below code can probably be cleaned up to be more generizable/less location dependent, but that is a project for another day.

```{r}

data <- read_excel("location/filename.xlsx", sheet = "sheetname")

```


#Data manipulation

First time using the *UpsetR* package. Attempting to do the manipulations necessary to make it work.

Subset/manipulate data to only have relevant variables. This is mostly needed because the Excel file isn't the best formatted.

```{r}

clean_data <- data %>%
  rename(Genes = "Gene name") %>%
  subset(select = c(Genes, Phenotypes)) %>%
  mutate(Phenotypes = str_replace_all(Phenotypes, ", ", "&"), 
         Phenotypes = str_replace_all(Phenotypes, " ", "_"), 
         Phenotypes = str_replace_all(Phenotypes, "pain", "Pain"), 
         Phenotypes = str_replace_all(Phenotypes, "Pain,", "Pain"))

```

Manipulating data to be in the format needed.

```{r}

upset_data <- clean_data %>%
  separate_longer_delim(Phenotypes, "&") %>%
  mutate(n = 1) %>%
  mutate(row = row_number()) %>%
  pivot_wider(names_from = Phenotypes, values_from = n, values_fill = 0) %>%
  select(-row) %>%
  group_by(Genes) %>%
  summarise(across(where(is.numeric), ~sum(.x))) %>%
  data.frame()

```


#Making plot

These actually turned out pretty well. 

**WARNING**: Put the library to load here instead of above because it will mess with *dplyr* otherwise.

##Setup

**WARNING**: Don't load the below package until the data manipulation is done! It will mess with *dplyr* so *group_by()* and *summarise()* won't work correctly.

```{r}

library(UpSetR)
library(ggplot2)

```


##Example

Example for *UpsetR* from Github. I found it useful to use this to get a sense of how different arguments work. The dataset *movies*, which it uses, isn't naturally present just be loading the library. Available from the Github I believe? I've just been using the actual code as a reference.

```{r}
upset(movies, 
      attribute.plots = list(gridrows = 100,
                             ncols = 1, 
                             plots = list(list
                                          (plot = histogram,
                                            x = "AvgRating",
                                            queries = T),
                                          list
                                          (plot = scatter_plot, 
                                            y = "AvgRating",
                                            x = "Watches",
                                            queries = T)
                                          )
                             ),
      sets = c("Action", "Adventure", "Children", "War", "Noir"),
      queries = list(list
                     (query = intersects, 
                       params = list("War"), 
                       active = T),
                     list(query = intersects, params = list("Noir")
                          )
                     )
      )
```


##Plotting

This is the code that actually makes the diagram. There are other parameters that can be added, but this makes the basic version.

```{r}

upset_plot <- upset(upset_data,
                    
      #this is the number of phenotypes/categories it will use. Make sure it is equal to the number present              
      nsets = 12, 
      
      #NA means all intersections are plotted. I suggest not messing with this one unless you're only trying to only look at specific intersections
      nintersects = NA, 
      
      #how large you want the points showing the intersections below the bar graph
      point.size = 1.5,
      
      #this makes it go from largest to smallest count
      order.by = "freq",
      
      #these are labels
      mainbar.y.label = "Phenotype Associations",
      sets.x.label = "Genes per Phenotype",
      
      #Sets the text size. Numbers indicate size of intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars, in that order
      text.scale = c(1.3, 1.3, 1, 1, 1, 1))

upset_plot

```



# Scatter plot

Attempting to make the phenotype by chromosome plot. This isn't necessarily relevant to anything, I just wanted to see if there were any interesting trends.

Starting with wrangling data into usable form.

```{r}

scatter_data <- data %>%
  rename(Chromosome = "Chromosome/scaffold name", Gene = "Gene name") %>%
  subset(select = c(Gene, Chromosome, Phenotypes)) %>%
  mutate(Phenotypes = str_replace_all(Phenotypes, ", ", "&"), 
         Phenotypes = str_replace_all(Phenotypes, " ", "_"), 
         Phenotypes = str_replace_all(Phenotypes, "pain", "Pain"), 
         Phenotypes = str_replace_all(Phenotypes, "Pain,", "Pain")) %>%
  separate_longer_delim(Phenotypes, "&") %>%
  mutate(Chromosome = fct_relevel(Chromosome, "1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15", "16", "17", "18", "19", "20", "21", "22", "X"))

```

Now generating plot.

```{r}

dist <- scatter_data %>%
  ggplot(aes(x = Phenotypes, y = Chromosome)) +
  geom_count(aes(color = after_stat(n), size = after_stat(n))) +
  guides(color = 'legend') +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

dist

```


#Exporting figure


```{r}

# The below code must be run all together. There may be an easier way to do this, but this is the best way I found to not cut off parts of the plot

# png(filename = "FILENAME.png") # or other device
# upset_plot
# dev.off()

```


